---
title: "Getting started with micstoich"
output: rmarkdown::html_vignette
vignette:
  %\VignetteIndexEntry{Getting started with micstoich}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
#library(micstoich)
devtools::load_all()
```

# Some simple examples

Let's use glucose oxidation as an example.

```{r}
micstoich(donor = 'C6H12O6', acceptor = 'O2')
```

Reactants have negative values and products positive.
The compounds themselves are given in the element names.

There is no biomass production in that example.
We can add it by providing a value for `fs`, perhaps 0.5 for this aerobic reaction.

```{r}
micstoich(donor = 'C6H12O6', acceptor = 'O2', fs = 0.5)
```

The default biomass composition is `C5H7O2N`, but it can be changed with `bioform`.

Is that reaction really balanced?
It *should* be, but we can check it or any reaction with `rxnbal()`.


```{r}
r1 <- micstoich(donor = 'C6H12O6', acceptor = 'O2', fs = 0.5)
r1
rxnbal(r1)
```

It returns `TRUE` invisibly if everything is OK.

```{r}
isTRUE(rxnbal(r1))
```

Here is an example that is not balanced.

```{r}
rxnbal(c(CH3COOH = -2, CH4 = 1, CO2 = 1))
```

We get a warning plus the overall balance, which shows 2 more moles of C and O and 4 of H on the reactant side than product side.
This is indicated by the negative values.

# Other electron acceptors

Let's use acetic acid as a donor with some other acceptors.

First nitrate for complete denitrification to N2.

```{r,error=TRUE}
micstoich(donor = 'CH3COOH', acceptor = 'NO3-')
```

Here we see we need to specify a product as well, because there are multiple possibilities.
We can find the one we want in the error message, where they are listed in `acceptor product` pairs: `'NO3- N2'`.
Or more easily, view them all with the `halfrxns()` function.

```{r}
halfrxns()
```

The one we want is in the third row.

```{r}
micstoich(donor = 'CH3COOH', acceptor = 'NO3-', product = 'N2')
```

How about sulfate?

```{r}
micstoich(donor = 'CH3COOH', acceptor = 'SO4-2', product = 'H2S')
```

But note that the package ignores elements other than C, H, O, and N in organic compounds, following Rittmann and McCarty (2001, 2020).
For example, if we have some empirical waste biomass formula like this one (spaces added for reading clarity, and could be removed),

```{r}
micstoich(donor = 'C226.36 H404.24 O187.69 N1 S0.52', acceptor = 'SO4-2', product = 'H2S')
```

we get an overall reaction, but it is not balanced.
And where is the problem?
We can check.

```{r}
r3 <- micstoich(donor = 'C226.36 H404.24 O187.69 N1 S0.52', acceptor = 'SO4-2', product = 'H2S')
rxnbal(r3)
```

The problem is extra S on the reactant side, in the substrate of course.
The negative sign in the results indicates that we have excess reactants, not products.
To use the package as intended, don't include S in the substrate!

```{r}
micstoich(donor = 'C226.36 H404.24 O187.69 N1', acceptor = 'SO4-2', product = 'H2S')
```

# Methanogenesis

Either use CO2 as the acceptor, but don't specify a product,

```{r}
micstoich(donor = 'CH3COOH', acceptor = 'CO2')
```

or specify the product, like this,

```{r}
micstoich(donor = 'CH3COOH', product = 'CH4')
```

Or both.

```{r}
micstoich(donor = 'CH3COOH', acceptor = 'CO2', product = 'CH4')
```

Hydrogenotrophic methanogenesis has similar flexibility.

```{r}
micstoich(donor = 'H2', acceptor = 'CO2')
micstoich(donor = 'H2', product = 'CH4')
micstoich(donor = 'H2', acceptor = 'CO2', product = 'CH4')
```

The different calls do result in different element order in the output.
We could use some of the other ordering options to align them.
The calls below demonstrate this, and have some cell biomass added to make the effects more clear.

```{r}
micstoich(donor = 'H2', acceptor = 'CO2', fs = 0.1, arrange = 'decreasing')
micstoich(donor = 'H2', acceptor = 'CO2', fs = 0.1, arrange = 'increasing')
micstoich(donor = 'H2', acceptor = 'CO2', fs = 0.1, arrange = 'alphanum')
```

All options list reactants first and products later.

# Fermentation

For fermentation, specify the organic product.

Glucose to acetic acid.

```{r}
micstoich(donor = 'C6H12O6', product = 'CH3COOH')
```

Or to ethanol.

```{r}
micstoich(donor = 'C6H12O6', product = 'CH3CH2OH')
```

Here we get carbon dioxide too.
This is why bread rises.

For production of only H2, use one of these approaches.

```{r}
micstoich(donor = 'C6H12O6', acceptor = 'H+')
micstoich(donor = 'C6H12O6', product = 'H2')
```

Of course, we have been ignoring cell biomass, but it could be added.

```{r}
micstoich(donor = 'C6H12O6', product = 'H2', fs = 0.2)
```

For mixed fermentations, just combine the products into one formula.
This example follows example 5.6 in Rittmann and McCarty (2020): fermentation of citrate to formate and acetate in a 1:2 molar ratio.
We need uncharged substrates, so will use citric acid.

```{r}
micstoich(donor = 'COOHCH2COHCOOHCH2COOH', product = 'HCOOH (CH3COOH)2')
```

# Other pathways

Mn reduction with organic donor
```{r}
micstoich('CH3COOH', acceptor = 'MnO2', product = 'Mn+2')
```

S reduction with H2 donor
```{r}
micstoich('H2', acceptor = 'S', product = 'H2S')
```

S reduction with organic donor
```{r}
micstoich('CH3COOH', acceptor = 'S', product = 'H2S')
```

Acetogenesis from H2
```{r}
micstoich('H2', acceptor = 'CO2', product = 'CH3COOH')
```

Chemolithotroph
```{r}
micstoich('Fe+2', acceptor = 'O2')
micstoich('NH3', acceptor = 'O2', product = 'N2')
micstoich('NH4+', acceptor = 'O2', product = 'N2')
```

Another inorganic donor
```{r}
micstoich('H2S', acceptor = 'O2', product = 'SO4-2')
```

## Bioremediation pathways

Arsenate reduction
```{r}
micstoich('CH3COOH', acceptor = 'HAsO4-2', product = 'H3AsO3')
```

Perchlorate reduction
```{r}
micstoich('CH3COOH', acceptor = 'ClO4-', product = 'Cl-')
```

Chromate reduction
```{r}
micstoich('CH3COOH', acceptor = 'CrO4-2', product = 'Cr+3')
```

Uranium reduction
```{r}
micstoich('CH3COOH', acceptor = 'UO2+2', product = 'U+4')
```

Selenate reduction
```{r}
micstoich('CH3COOH', acceptor = 'SeO4-2', product = 'Se')
```

## Fumarate reduction

Fumarate reduction (fumaric acid = C4H4O4, succinic acid = C4H6O4)
```{r}
micstoich('CH3COOH', acceptor = 'C4H4O4', product = 'C4H6O4')
```

# Mixed substrates
Mixed substrates follow the mixed product example.

For example, a mix of lactose and protein (skim milk?) fermented to a mix of lactic and acetic acids plus hydrogen:

```{r}
micstoich(donor = '(C12H22O11)5 (C5H7O2N)', product = '(C3H6O3)3 CH3COOH (H2)5')
```

The mixed donor and product are no problem, but we have to manually separate the results.

# Errors in `micstoich()`

The following calls are missing something, and the error messages are meant to explain the problem.

```{r, error = TRUE}
micstoich(donor = 'C6H12O6')
micstoich(donor = 'C6H12O6', acceptor = 'Fe')
micstoich(donor = 'C100P', acceptor = 'O2')
```

Presently mistakes in element names can give confusing errors.

```{r, error = TRUE}
micstoich(donor = 'X', acceptor = 'O2')
```

A good way to proceed in case of a strange error message is to try the lower-level functions described below.
For example:

```{r, error = TRUE}
molmass('X')
calcCOD('X')
```

The next error is related to how all compounds are identified internally--simply by the character strings provided by formulas (or calculated).

```{r, error = TRUE}
micstoich(donor = 'C5H6O2N', product = 'C5H6O2N')
micstoich(donor = 'C5H7O2N', acceptor = 'O2', bioform = 'C5H7O2N', fs = 0.2)
```

They can be fixed by simply changing element order,

```{r}
micstoich(donor = 'C5H6O2N', product = 'H6C5O2N')
micstoich(donor = 'H7C5O2N', acceptor = 'O2', bioform = 'C5H7O2N', fs = 0.2)
```

or even just adding a space.

```{r}
micstoich(donor = 'C5 H6O2N', product = 'C5H6O2N')
micstoich(donor = 'C5 H7O2N', acceptor = 'O2', bioform = 'C5H7O2N', fs = 0.2)
```

# Molar mass and COD'

For convenience, the package includes a function for calculating molar mass and "calculated" or "theoretical" chemical oxygen demand, COD'.
These are closely related to the functions with similar names in the biogas package.
Any of the chemical formulas used above work here.

```{r}
molmass('C6H12O6')
molmass('C226.36 H404.24 O187.69 N1 S0.52')
```

And so do others.

```{r}
molmass('S')
molmass('NaCl')
molmass('XePtF6')
```

COD' is returned as g oxygen per g of the given compound by default.

```{r}
calcCOD('C6H12O6')
calcCOD('C226.36 H404.24 O187.69 N1 S0.52')
```

But this could be changed to per mole.

```{r}
calcCOD('C6H12O6', per = 'mol')
calcCOD('C226.36 H404.24 O187.69 N1 S0.52', per = 'mol')
```

As expected, COD' will be zero for inorganic compounds.

```{r}
calcCOD('NaCl')
```

# More information
For more on the micstoich package, see <https://github.com/AU-BCE-EE/micstoich>.

# References
Rittmann, B.E. and McCarty, P.L. (2001) *Environmental Biotechnology:
Principles and Applications*. McGraw-Hill, New York.

Rittmann, B.E. and McCarty, P.L. (2020) *Environmental Biotechnology:
Principles and Applications*. 2nd ed. McGraw-Hill, New York.

