---
title: "Getting started with micstoich"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting started with micstoich}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
library(micstoich)
source('../R/core.R')
source('../R/halfrxn.R')
source('../R/constants.R')
```

# Some simple examples

Let's use glucose oxidation as an example.

```{r}
micstoich(donor = 'C6H12O6', acceptor = 'O2')
```

Reactants have negative values and products positive.
The compounds themselves are given in the element names.

There is no biomass production in that example.
We can add it by providing a value for `fs`, perhaps 0.5.

```{r}
micstoich(donor = 'C6H12O6', acceptor = 'O2', fs = 0.5)
```

The default biomass composition is `C5H7O2N`, but it can be changed with `bioform`.

Is it really balanced?
We can check the output or any reaction with `rxnbal()`.


```{r}
r1 <- micstoich(donor = 'C6H12O6', acceptor = 'O2', fs = 0.5)
r1
rxnbal(r1)
isTRUE(rxnbal(r1))
```

It returns `TRUE` invisibly if everything is OK.

# Other electron acceptors

Let's use acetic acid as a donor with some other acceptors.

First nitrate for complete denitrification to N2.

```{r,error=TRUE}
micstoich(donor = 'CH3COOH', acceptor = 'NO3-')
```

Here we see we need to specify a product as well, because there are multiple possibilities.
We can find the one we want (`'NO3- N2'`) in the error message, or more easily, view them all with the `halfrxns()` function.

```{r} 
halfrxns()
```

Indeed, the one we want is in the third row.

```{r,error=TRUE}
micstoich(donor = 'CH3COOH', acceptor = 'NO3-', product = 'N2')
```

How about sulfate?

```{r}
micstoich(donor = 'CH3COOH', acceptor = 'SO4-2', product = 'H2S')
```

# Methanogenesis

Either use CO2 as the acceptor, but don't specify a product.

```{r}
micstoich(donor = 'CH3COOH', acceptor = 'CO2')
```

Or specify the product.

```{r}
micstoich(donor = 'CH3COOH', product = 'CH4')
```

Or both.

```{r}
micstoich(donor = 'CH3COOH', acceptor = 'CO2', product = 'CH4')
```

# Fermentation

For fermentation, specify the organic product.

Glucose to acetic acid.

```{r}
micstoich(donor = 'C6H12O6', product = 'CH3COOH')
```

Or to ethanol.

```{r}
micstoich(donor = 'C6H12O6', product = 'CH3CH2OH')
```

This is why bread rises.

For production of only H2, use these approaches.

```{r}
micstoich(donor = 'C6H12O6', acceptor = 'H+')
micstoich(donor = 'C6H12O6', product = 'H2')
```

Of course, we have been ignoring cell biomass, but it could be added.

```{r}
micstoich(donor = 'C6H12O6', product = 'H2', fs = 0.2)
```

For mixed fermenations, just combine the products into one formula.
This example follows example 5.6 in Rittmann and McCarty (2020): fermentation of citrate to formate and acetate in a 1:2 molar ratio.
We need uncharged substrates, so will use citric acid.

```{r}
micstoich(donor = 'COOHCH2COHCOOHCH2COOH', product = 'HCOOH (CH3COOH)2')
```

# Mixed substrates
Mixed substrates follow the mixed product example.

For example, a mix of lactose and protein (skim milk?) fermented to a mix of lactic and acetic acids plus hydrogen:

```{r}
micstoich(donor = '(C12H22O11)2 (C5H7O2N)3', product = '(C3H6O3)3 CH3COOH (H2)5')
```




# With biomass
micstoich(donor = 'C6H12O6', product = 'CH3CH2OH', bioform = 'C5H7O2N', fs = 0.1)
# H2
micstoich(donor = 'C6H12O6', acceptor = 'H+')
# Mixed products
micstoich(donor = 'C6H12O6', product = 'CH3COOH CH2CH2COH (H2)3')
micstoich(donor = 'C6H12O6', product = 'H2', bioform = 'C5H7O2N', fs = 0.1)
```

# Mixed reactants

```{r}
micstoich(donor = 'CH3COOH H2', acceptor = 'O2')
```

# Anaerobic respiration with multiple possible products

In some cases the same acceptor can produce different products.
Use the `product` argument to specify which one.

```{r}
micstoich(donor = 'C6H12O6', acceptor = 'NO3-')
micstoich(donor = 'C6H12O6', acceptor = 'NO3-', product = 'N2')
micstoich(donor = 'C6H12O6', acceptor = 'NO3-', product = 'NH3')
micstoich(donor = 'C6H12O6', acceptor = 'NO3-', product = 'NO2-')
```

# Errors

```{r, error = TRUE}
micstoich(donor = 'C6H12O6')
micstoich(donor = 'C6H12O6', acceptor = 'Fe')
micstoich(donor = 'C100P', acceptor = 'O2')
orgrxn('HCOOH') * .111
orgrxn('CH3COOH') * .889
orgrxn('HCOOH (CH3COOH)2')
```
